\documentclass[12pt]{letter}
\usepackage[margin=1in]{geometry}

\signature{Andrew H. Bond\\Senior Member, IEEE\\
Department of Computer Engineering\\
San Jose State University}

\address{Department of Computer Engineering\\
San Jose State University\\
San Jose, CA 95192, USA\\
andrew.bond@sjsu.edu}

\begin{document}

\begin{letter}{Editor-in-Chief\\
IEEE Transactions on Artificial Intelligence}

\opening{Dear Editor,}

I am pleased to submit the enclosed manuscript, ``Formula Ceiling
Detection: Phased Enumeration for Interpretable Symbolic Prediction,''
for consideration as a Regular Paper in IEEE Transactions on
Artificial Intelligence.

The manuscript introduces a method for discovering interpretable
predictive formulas from scientific data by exhaustively enumerating
compositions of mathematical operations in phases of increasing
complexity. The method's primary contribution is the \emph{formula
ceiling}: the highest prediction score (F1) achievable by any
enumerated formula. Comparing this ceiling to a black-box ensemble
yields the \emph{complexity gap}---a diagnostic that tells
practitioners whether their prediction task admits a simple formula
or genuinely requires a complex model.

We validate the method on five classification benchmarks (including
the UCI Breast Cancer and Ionosphere datasets) and compare directly
against PySR, a leading genetic symbolic regression method. On
classification tasks, our method matches or exceeds PySR while
additionally providing the ceiling diagnostic. We further demonstrate
the method on a scientific discovery task---predicting periodic orbits
in the gravitational three-body problem---where it discovers a
two-feature formula and identifies a complexity gap of 0.35 to
gradient boosting.

The method is released as an open-source Python package
(\texttt{pip install symbolic-search}) with 23 unit tests and
continuous integration, and all experimental data is publicly
available.

This manuscript has not been published elsewhere and is not under
consideration by another journal. All work is original.

\closing{Sincerely,}

\end{letter}
\end{document}
