\documentclass[journal]{IEEEtran}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{url}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{cite}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

\begin{document}

\title{Formula Ceiling Detection: Phased Enumeration for Interpretable Symbolic Prediction}

\author{Andrew~H.~Bond,~\IEEEmembership{Senior Member,~IEEE}
\thanks{A. H. Bond is with the Department of Computer Engineering,
San Jose State University, San Jose, CA 95192 USA
(e-mail: andrew.bond@sjsu.edu). ORCID: 0009-0003-2599-6158.}
}

\markboth{IEEE Transactions on Artificial Intelligence}
{Bond: Formula Ceiling Detection}

\maketitle

\begin{abstract}
We introduce a framework for discovering interpretable predictive
formulas from labeled scientific data by enumerating compositions of
mathematical operations in phases of increasing complexity. Given a
feature matrix and binary labels, the method evaluates all single
features, all pairwise operation combinations, and selected higher-order
compositions, finding the formula that maximizes F1 score within each
phase. GPU-vectorized evaluation enables thousands of candidates per
second.

The framework's primary contribution is the \emph{formula ceiling}:
the highest F1 achieved by any enumerated formula. Comparing this
ceiling to a black-box ensemble (trained and evaluated on the same
data split) yields the \emph{complexity gap}---how much predictive
structure lies beyond the enumerated formula space.

Applied to predicting periodic orbits in the gravitational three-body
problem (15{,}000 configurations, 23 features, 5-fold cross-validated),
the method discovers that periodicity is predicted by
$\sqrt{n_\mathrm{clusters}^2 + \lambda_\mathrm{min}^2}$
(ceiling F1~$= 0.48$). A gradient-boosted ensemble on the same data
and splits reaches F1~$= 0.83$, yielding a complexity gap of 0.35.
Across five classification benchmarks, the ceiling correctly
identifies problems admitting exact formulas (Circles: ceiling
$= 0.996$, beating the ensemble) and problems requiring complex
models (three-body: gap $= 0.35$).

Code and data: \url{https://github.com/ahb-sjsu/astar-symbolic-discovery}.
\end{abstract}

\section{Introduction}

A central goal of computational science is to discover interpretable
laws from data. This problem has been addressed by genetic symbolic
regression~\cite{schmidt2009distilling, cranmer2023interpretable},
neural equation discovery~\cite{udrescu2020ai}, and classical AI
systems~\cite{langley1987scientific}. Genetic programming explores
large formula spaces but provides no optimality guarantee at a given
depth and often produces bloated expressions. Neural methods discover
accurate but opaque representations. Classical systems require
domain-specific search heuristics.

We propose \emph{phased enumeration}: exhaustive evaluation of all
formulas within a series of structural templates of increasing
complexity. Phase~1 evaluates all single features; Phase~2 evaluates
all ordered pairs of features composed with each binary operation;
Phase~3 applies unary transforms to the top Phase~2 results.
Within Phases~1--2, every candidate is evaluated, so the best formula
\emph{within that phase's template} is found by complete enumeration.
Phase~3 is restricted to the top-$K$ Phase~2 results and is therefore
not exhaustive over all depth-3 formulas. Formulas with structures
not matching any template (e.g., multi-term sums, deeply nested
compositions) are not searched.

Our framework's primary contribution is the \emph{formula ceiling}:
\begin{enumerate}
\item The \textbf{ceiling} at each phase: the highest F1 achieved by
any enumerated formula. If the ceiling stabilizes across phases, the
search has likely saturated for the given operation set.
\item The \textbf{complexity gap}: the difference between a black-box
ensemble's F1 and the ceiling, both evaluated on the same data using
the same cross-validation splits. A positive gap indicates predictive
structure beyond the enumerated formula space; a zero or negative gap
indicates a simple formula suffices.
\end{enumerate}

\section{Method}

\subsection{Problem Formulation}

\begin{definition}[Symbolic Prediction Problem]
Given a feature matrix $X \in \mathbb{R}^{N \times d}$ and binary
labels $y \in \{0, 1\}^N$, find a formula
$f: \mathbb{R}^d \to \mathbb{R}$ and threshold $\tau \in \mathbb{R}$
that maximize $F_1(\hat{y}, y)$ where
$\hat{y}_i = \mathbf{1}[f(X_i) > \tau]$, subject to
$\mathrm{depth}(f) \leq D$.
\end{definition}

\subsection{Operation Registry}

The search space is defined by three operation classes:

\textbf{Leaf nodes} (depth 0): Each of the $d$ features $x_j$.

\textbf{Unary operations} (depth $+1$): The full registry contains 8
operations: $\log|\cdot|$, $\sqrt{|\cdot|}$, $(\cdot)^2$, $1/(\cdot)$,
$-(\cdot)$, $|\cdot|$, $\sigma(\cdot)$, $\tanh(\cdot)$.
In the experiments reported here, we use a subset of 5:
$\{\log, \sqrt{}, (\cdot)^2, 1/(\cdot), -(\cdot)\}$.

\textbf{Binary operations} (depth $+1$): The full registry contains 10
operations. In our experiments we use 7:
$\{+, -, \times, \div, \max, \min, \mathrm{hypot}\}$.
The remaining 3 (diff-squared, harmonic mean, geometric mean) are
available in the software but were not used in the reported runs.

\subsection{Phased Exhaustive Search}

Rather than a single tree search, we organize the exploration into
four phases (Algorithm~\ref{alg:search}). Phases~1--2 enumerate all
formulas of their structural type; Phases~3--4 are restricted to
top-$K$ results from earlier phases and are not exhaustive.

\begin{algorithm}[t]
\caption{Phased Symbolic Search (Phases~1--2 exhaustive; 3--4 filtered)}
\label{alg:search}
\begin{algorithmic}[1]
\Require Features $X \in \mathbb{R}^{N \times d}$, labels $y$,
operations $\mathcal{U}$ (unary), $\mathcal{B}$ (binary)
\Ensure Best formula and F1 at each depth
\Statex
\Statex \textbf{Phase 1:} Single features (depth 1)
\For{$j = 1, \ldots, d$}
    \State Evaluate $F_1^*(x_j)$ via threshold sweep
\EndFor
\State $R_1 \gets$ top-$K$ features by F1
\Statex
\Statex \textbf{Phase 2:} Pairwise combinations (depth 2)
\For{$x_i, x_j \in \{1,\ldots,d\}$, $b \in \mathcal{B}$}
    \State Evaluate $F_1^*(b(x_i, x_j))$ via threshold sweep
\EndFor
\State $R_2 \gets$ top-$K$ pairwise formulas
\Statex
\Statex \textbf{Phase 3:} Unary transforms (depth 3)
\For{$f \in R_2$, $u \in \mathcal{U}$}
    \State Evaluate $F_1^*(u(f))$ via threshold sweep
\EndFor
\State $R_3 \gets$ top-$K$ depth-3 formulas
\Statex
\Statex \textbf{Phase 4:} Triple combinations (depth 3)
\For{$x_i, x_j, x_k \in R_1$, pattern $p$}
    \State Evaluate $F_1^*(p(x_i, x_j, x_k))$ via threshold sweep
\EndFor
\State \Return best formula across all phases
\end{algorithmic}
\end{algorithm}

\textbf{Threshold sweep.} For each candidate formula $f$, we compute
$v_i = f(X_i)$ for all $N$ samples, then evaluate F1 at 11 percentile
thresholds $\{20, 30, 40, \ldots, 95\}$ of the value distribution.
Both $v > \tau$ and $v < \tau$ predictors are tested. The best
threshold is retained. This discrete sweep approximates but does not
guarantee the globally optimal threshold; in practice, F1 is
piecewise-constant in $\tau$ (changing only when $\tau$ crosses a
data point), and our 11 percentiles sample this step function densely
enough that the reported F1 matches exhaustive threshold search on
a verified subset (median absolute error $< 0.001$).
The sweep is vectorized and adds negligible cost.

\textbf{Cross-validation protocol.} To ensure a fair comparison
between the formula ceiling and the ensemble baseline, both are
evaluated using the same 5-fold stratified split. For each fold:
(i)~the phased search runs on the training portion (formula selection
and threshold tuning); (ii)~the best formula + threshold is evaluated
on the held-out test portion; (iii)~the gradient-boosted ensemble is
trained and evaluated on the same train/test split. The reported
ceiling and ensemble F1 are both averages over the 5 held-out folds.
This prevents information leakage from threshold selection into the
ceiling estimate.

\textbf{GPU acceleration.} Features are stored as GPU arrays (CuPy).
All formula evaluations and threshold sweeps are vectorized GPU
operations. Phase~2 evaluates $d(d-1) \times |\mathcal{B}|$ formulas
(e.g., $23 \times 22 \times 7 = 3{,}542$ with our 7-operation set)
in seconds.

\textbf{Completeness within each phase.} Phases~1--2 are exhaustive
over all feature (pairs) and operations. Phase~3 applies all unary
transforms to the top-$K$ results from Phase~2. Phase~4 tests
multiple combination patterns over the top-$K$ single features.
The search is complete at each phase for the specified formula
structure; no formulas of that structure are missed.

\subsection{Scalability}

The formula count per phase is:
Phase~1: $d(1 {+} |\mathcal{U}|)$;
Phase~2: $d(d{-}1)|\mathcal{B}|$;
Phase~3: $K|\mathcal{U}|$;
Phase~4: $\binom{K'}{3} P$.
Here $K$ is the number of top pairwise results carried forward,
$K'$ the top single features, and $P$ the combination patterns.

For $d = 23$, $|\mathcal{B}| = 7$, $|\mathcal{U}| = 5$, $K = 30$,
$K' = 8$, $P = 4$, the total is approximately 5{,}600 formulas---evaluated
in under one minute on a single GPU. Composing Phase~2 with itself
(all pairwise-of-pairwise formulas) would require
$O(d^2(d-1)^2 \cdot |\mathcal{B}|^2) \approx 12$ million formulas,
tractable with GPU vectorization but requiring hours.
We report results through Phase~3 (depth 3); Phase~4 triples are
also depth~3 but with a different structural template.

\subsection{Formula Ceiling and Irreducible Complexity Gap}

\begin{definition}[Formula Ceiling]
The formula ceiling $C_D$ at depth $D$ is the maximum F1 achieved
by any formula of depth $\leq D$ found across all phases:
$C_D = \max_{\text{phases}} F_1^*$.
\end{definition}

\begin{definition}[Irreducible Complexity Gap]
Given a black-box model with F1 score $F_1^{\mathrm{bb}}$ trained on
the same features, the irreducible complexity gap at depth $D$ is
$G_D = F_1^{\mathrm{bb}} - C_D$.
\end{definition}

$G_D > 0$ means the data contains predictive structure that was not
captured by any depth-$D$ formula within the searched space. Because
the search is exhaustive within each phase template but does not
cover all possible depth-$D$ formulas, $G_D$ is an upper bound on
the true gap---a formula structure not in our template set could in
principle achieve higher F1. However, if $C_D$ is stable across
multiple operation sets and phase structures, this suggests (but
does not prove) that the gap is intrinsic to the depth constraint
rather than a limitation of the search.

\begin{proposition}
$C_D$ is non-decreasing in $D$ and bounded above by
$F_1^{\mathrm{Bayes}}$. If $C_{D+1} = C_D$ for a sufficiently
rich operation set, the ceiling has converged.
\end{proposition}

\section{Benchmark Validation}

To validate the method beyond the three-body case study, we run the
phased search on four standard classification datasets and compare
the formula ceiling to gradient-boosted ensembles (Table~\ref{tab:bench}).

\begin{table}[h]
\caption{Formula ceiling across domains. Both ceiling and GB use
5-fold CV: formula selection and threshold tuning occur on training
folds only, with F1 measured on each held-out fold.
Gap $= \text{GB} - C$.}
\label{tab:bench}
\scriptsize
\begin{tabular}{@{}lccrl@{}}
\toprule
Dataset ($d$) & Ceiling & GB F1 (CV) & Gap & Formula \\
\midrule
Circles (2) & 0.996 & $0.989{\pm}.01$ & $-$0.01 & \texttt{hypot} \\
Breast Ca.\ (30) & 0.948 & $0.962{\pm}.01$ & +0.01 & \texttt{f7+f22} \\
Ionosphere (34) & 0.902 & $0.934{\pm}.03$ & +0.03 & \texttt{log(f6 min f5)} \\
Moons (2) & 0.833 & $0.969{\pm}.01$ & +0.14 & \texttt{x2} \\
Three-body (23) & 0.480 & $0.830{\pm}.02$ & +0.35 & \texttt{hypot(n,\,$\lambda$)} \\
\bottomrule
\end{tabular}
\end{table}

On \textbf{Circles}, the method discovers the exact analytic decision
boundary $\sqrt{x_1^2 + x_2^2} > \tau$ and \emph{outperforms} gradient
boosting ($F_1 = 0.996$ vs.\ $0.989$). On \textbf{Breast Cancer} and
\textbf{Ionosphere}, simple two-feature formulas achieve 95--98\% of
ensemble performance. On \textbf{Moons}, the nonlinear boundary resists
depth-3 formulas (gap $= 0.136$). On the \textbf{Three-body} problem,
the gap is largest ($0.35$), indicating substantial nonlinear structure.

On these benchmarks, the ceiling distinguishes problems where simple
formulas suffice (Circles: gap $< 0$), problems where formulas are
close (Breast Cancer: gap $\approx 1\%$), and problems where the
enumerated formulas leave substantial room for improvement
(Three-body: gap $= 35\%$). All gaps use ensemble and ceiling
on the same data and CV splits.

\section{Case Study: Three-Body Periodic Orbit Prediction}

\subsection{Setup}

We use a single dataset of 15{,}000 random configurations of
equal-mass bodies ($m_1 = m_2 = m_3 = 1$). Both the formula search
and the gradient-boosted baseline are evaluated on this dataset using
the same 5-fold cross-validation splits, ensuring the ceiling and
ensemble F1 are directly comparable.

Features include: numerical matrix rank (threshold $\epsilon = 10^{-6}$),
potential block eigenvalues and their ratios, frequency cluster counts,
spectral entropy, Hamiltonian energy, body separation ratios, and
derived quantities. Ground truth (periodic vs.\ non-periodic) is
obtained by symplectic leapfrog integration ($\Delta t = 0.005$,
$8 \times 10^4$ steps, $T = 400$ natural time units) on an NVIDIA
GV100 GPU, with periodicity defined as phase-space return to within
5\% of initial extent after the first 10\% of integration time.
Batch size is maximized automatically using the batch-probe
library~\cite{batchprobe}.

\subsection{Search Results}

Table~\ref{tab:phases} summarizes the results across all four phases.

\begin{table}[h]
\caption{Formula search results by phase (15{,}000 configs, 23 features).
All phases search formulas of depth $\leq 3$.}
\label{tab:phases}
\scriptsize
\begin{tabular}{@{}llrr@{}}
\toprule
Phase (depth) & Best formula & F1 & Tested \\
\midrule
1: single+unary (1--2) & \texttt{n\_clusters} & 0.35 & 138 \\
2: pairwise (2) & \texttt{n\_clust + $\lambda_\mathrm{min}$} & 0.48 & 3{,}542 \\
3: unary of top-30 (3) & \texttt{hypot(n\_cl, $\lambda_\mathrm{min}$)} & 0.48 & 150 \\
4: triple of top-8 (3) & (none above 0.30) & --- & 224 \\
\midrule
\textbf{Best (depth $\leq$ 3)} & & \textbf{0.48} & \textbf{4{,}054} \\
\bottomrule
\end{tabular}
\end{table}

The best F1 is reached at Phase~2 and does not improve in Phases~3--4.
All 20 top formulas are algebraically equivalent to linear combinations
of $n_\mathrm{clusters}$ and $\lambda_\mathrm{min}$. The Euclidean
norm variant
\begin{equation}
\hat{y} = \mathbf{1}\!\left[
\sqrt{n_\mathrm{clusters}^2 + \lambda_\mathrm{min}^2} > \tau
\right]
\end{equation}
is representative of the equivalence class.

\subsection{Irreducible Complexity Gap}

\begin{table}[h]
\caption{Formula ceiling vs.\ black-box models (all on 15K dataset,
5-fold CV for ensemble methods; $\pm$ shows fold standard deviation).}
\label{tab:gap}
\begin{tabular}{lcc}
\toprule
Model & F1 & Gap \\
\midrule
Random (4\% base rate) & $\sim$0.07 & --- \\
$r_2/r_1 > 10$ & 0.29 & --- \\
Best enumerated formula & 0.48 & 0.00 \\
Decision tree (CV) & $0.55 \pm 0.03$ & $-$0.07 \\
Gradient boosting (CV) & $0.83 \pm 0.02$ & $-$0.35 \\
\bottomrule
\end{tabular}
\end{table}

The observed gap is $G = 0.83 - 0.48 = 0.35$ (both measured on the
same 15K dataset with 5-fold CV for the ensemble). The decision tree
(F1~$= 0.55$, CV) captures some of this gap via piecewise-constant
splits; the remaining $0.83 - 0.55 = 0.28$ requires the nonlinear
interactions of an ensemble. Within the enumerated formula structures,
no expression achieves more than 58\% of the ensemble's predictive
power on this dataset. This gap may narrow with deeper search, richer
operations, or additional features, but its stability across Phases~2--3
and across all tested operation types is consistent with genuine
data complexity, though a richer formula space (e.g., multi-term
sums, deeper nesting) could in principle narrow the gap.

\subsection{Physical Interpretation}

The discovered formula combines two independent necessary conditions:
\begin{itemize}
\item $n_\mathrm{clusters} \geq 2$: the eigenvalue frequency spectrum
separates into fast (inner binary) and slow (outer body) groups,
indicating timescale separation.
\item $\lambda_\mathrm{min} > 0$: the smallest Hessian eigenvalue is
nonzero, indicating the system is gravitationally bound (not
escaping).
\end{itemize}
The Euclidean norm indicates these contribute independently---there
is no synergy or interaction between them. Periodicity requires
\emph{both} separated timescales \emph{and} sufficient binding.

\subsection{Comparison to PySR}

We compare directly against PySR~\cite{cranmer2023interpretable} on
four AI Feynman regression problems (Table~\ref{tab:pysr}).

\begin{table}[h]
\caption{Regression: our method vs.\ PySR (R$^2$, 3{,}000 samples).}
\label{tab:pysr}
\scriptsize
\begin{tabular}{@{}lcccl@{}}
\toprule
Problem & Ours & PySR & PySR time & PySR formula \\
\midrule
Gravity & 0.99 & \textbf{1.00} & 21s & \texttt{m1*m2/r\^{}2} (exact) \\
Coulomb & 0.99 & \textbf{1.00} & 5s & \texttt{q1*q2/r\^{}2} (exact) \\
$x_1^2{+}x_2^2$ & 0.50 & \textbf{1.00} & 6s & \texttt{sq(x1)+sq(x2)} (exact) \\
Thin lens & 0.93 & 0.96 & 7s & approximate \\
\bottomrule
\end{tabular}
\end{table}

On regression (Table~\ref{tab:pysr}), PySR recovers exact formulas on
3/4 problems by exploring deeper trees (depth~10+). Our method matches
on Gravity and Coulomb (depth~2 suffices) but misses $x_1^2 + x_2^2$
(requires multi-term addition outside our Phase~2 template).

For a direct apples-to-apples comparison on the paper's classification
task, we also ran PySR on the same classification datasets, fitting
binary labels as continuous regression targets and thresholding the
output for F1 (Table~\ref{tab:pysr_class}).

\begin{table}[h]
\caption{Classification F1: our method vs.\ PySR (same datasets,
same F1 metric via threshold sweep on PySR's continuous output).}
\label{tab:pysr_class}
\scriptsize
\begin{tabular}{@{}lccc@{}}
\toprule
Dataset & Ours & PySR & PySR formula \\
\midrule
Circles & \textbf{0.996} & 0.996 & \texttt{sq(sq(x2*x2)...)} \\
Moons & 0.833 & 0.834 & \texttt{log(2.05 - x2)} \\
Breast Ca. & \textbf{0.948} & 0.927 & \texttt{sqrt(exp(...))} \\
\bottomrule
\end{tabular}
\end{table}

Our method matches PySR on Circles (both find the radial boundary)
and Moons (both limited by the nonlinear boundary), and outperforms
PySR on Breast Cancer ($0.948$ vs.\ $0.927$). PySR's deeper trees do
not help on these datasets because the optimal formulas are shallow.
The key distinction remains: our framework additionally provides the
ceiling diagnostic and complexity gap.

\subsection{Convergence and Ablation}

Table~\ref{tab:convergence} shows the ceiling at each search phase.

\begin{table}[h]
\caption{Ceiling convergence: F1 at each phase.}
\label{tab:convergence}
\small
\begin{tabular}{lccc}
\toprule
Dataset & Phase 1 & Phase 2 & Phase 3 \\
\midrule
Circles & 0.730 & \textbf{0.996} & 0.996 \\
Moons & \textbf{0.833} & 0.833 & 0.833 \\
Breast Cancer & 0.926 & \textbf{0.934} & 0.934 \\
\bottomrule
\end{tabular}
\end{table}

The ceiling converges at Phase~2 for all datasets. Phase~3 (unary
transforms) does not improve the best score, suggesting that the ceiling
has stabilized for the tested operations.

\textbf{Ablation.} On Circles, removing the \texttt{hypot} operation
drops the ceiling from 0.996 to 0.796. All other operations are
redundant---\texttt{hypot} alone captures the circular decision
boundary. This demonstrates that the ablation identifies which
operations are structurally necessary.

\textbf{Fold stability.} Table~\ref{tab:folds} shows per-fold
formula selection across the 5-fold CV. On Circles, the same formula
(\texttt{x1 hypot x2}) is selected on all 5 folds with negligible
F1 variance ($\pm 0.001$). On Moons, \texttt{x2} is selected
unanimously. On Breast Cancer, two formula families alternate
(\texttt{f0+f7} and \texttt{f3 max f7}), reflecting the dataset's
multiple near-equivalent feature pairs. In all cases, the ceiling
is stable ($\sigma \leq 0.004$).

\begin{table}[h]
\caption{Per-fold ceiling stability (5-fold CV).}
\label{tab:folds}
\scriptsize
\begin{tabular}{@{}lccl@{}}
\toprule
Dataset & Ceiling ($\mu \pm \sigma$) & GB ($\mu \pm \sigma$) & Formula (folds) \\
\midrule
Circles & $0.996 \pm 0.001$ & $0.985 \pm 0.007$ & \texttt{hypot} (5/5) \\
Moons & $0.833 \pm 0.004$ & $0.966 \pm 0.008$ & \texttt{x2} (5/5) \\
Breast Ca. & $0.937 \pm 0.004$ & $0.950 \pm 0.019$ & \texttt{f0+f7} (2/5) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Runtime Scaling}

\begin{table}[h]
\caption{Runtime vs.\ feature count (Phase~1+2, single CPU core).}
\label{tab:scaling}
\small
\begin{tabular}{rrrr}
\toprule
$d$ & Formulas & Time (s) & Best F1 \\
\midrule
2 & 16 & 0.01 & 0.96 \\
5 & 145 & 0.09 & 0.98 \\
10 & 640 & 0.38 & 0.96 \\
15 & 1{,}485 & 0.91 & 0.97 \\
20 & 2{,}680 & 1.66 & 0.95 \\
\bottomrule
\end{tabular}
\end{table}

Runtime scales as $O(d^2)$, reaching 1.7~seconds at $d = 20$.
GPU vectorization (not used here) would further reduce these times.
At $d = 50$, the estimated Phase~2 time is $\sim$20~seconds on CPU.

\section{Discussion}

\subsection{When to Use This Method}

The phased exhaustive approach is most valuable when:
(i)~the feature count is moderate ($d < 50$, so $d^2$ is tractable);
(ii)~the practitioner needs the best formula within a bounded depth,
not a stochastic approximation;
(iii)~the formula ceiling and irreducible complexity gap are of
interest (e.g., to decide whether a problem admits a simple law or
requires a complex model).

For large feature spaces ($d > 100$) or deep formula trees
(depth $> 5$), genetic methods (PySR) or neural symbolic regression
(AI Feynman) are more appropriate. Our method's advantage is the
ceiling diagnostic: PySR finds better formulas on deep problems,
but does not quantify when further search is futile.

\subsection{The Ceiling as a Scientific Diagnostic}

The formula ceiling is a statement about the interaction between the
data and the searched formula space. A converged ceiling with a
positive gap means that the enumerated formulas do not capture the
full predictive structure available to flexible models. This could
indicate genuine nonlinearity in the data, or it could indicate that
the searched templates miss a formula that would close the gap. In
our three-body case study, the ceiling of 0.48 is stable across
Phases~2--3 and across all 7 binary and 5 unary operations tested,
which is consistent with---but does not prove---genuine complexity
beyond the depth-3 formula space.

This diagnostic is novel. Existing symbolic regression methods report
the best formula found but do not quantify \emph{how much better}
a more complex model could do. The ceiling provides a principled
stopping criterion: if $C_{D+1} = C_D$ and $G_D > 0$, stop searching
for formulas and use an ensemble.

\subsection{Limitations}

The search is exhaustive only within each phase's structure.
Phase~2 (pairwise) is complete; Phase~3 applies unary transforms
only to top-$K$ pairwise results, not all; Phase~4 tests only
pre-specified triple patterns. A formula with structure not matching
any phase template (e.g., $\log(x_1 + x_2) \cdot x_3 / x_4$) will
be missed.

The method does not discover features---it evaluates pre-computed
ones. The quality of the discovered formula depends entirely on the
feature engineering.

The F1 metric weights precision and recall equally. Domain-specific
objectives (e.g., prioritizing recall for safety applications) could
change the optimal formula.

\section{Conclusion}

We have introduced phased enumeration with formula ceiling detection
for symbolic prediction. The method enumerates all formulas within
defined structural templates and measures the complexity gap to a
black-box ensemble on the same data split. Applied to the three-body
problem, the ceiling stabilizes at F1~$= 0.48$ while gradient
boosting reaches $0.83$ on the same dataset, yielding a gap of $0.35$.
Across five classification benchmarks, the ceiling distinguishes
problems where formulas match ensemble performance (Circles) from
problems where a substantial gap remains (three-body, Moons).

The method is domain-general, requiring only a feature matrix and
binary labels. It is practical for feature spaces up to $d \approx 50$
and formula depths up to 3, with GPU acceleration enabling evaluation
of $\sim$6{,}000 formulas in under one minute.

\begin{thebibliography}{9}

\bibitem{schmidt2009distilling}
M.~Schmidt and H.~Lipson, ``Distilling free-form natural laws from
experimental data,'' \emph{Science}, vol.~324, pp.~81--85, 2009.

\bibitem{udrescu2020ai}
S.-M.~Udrescu and M.~Tegmark, ``AI Feynman: A physics-inspired
method for symbolic regression,'' \emph{Science Advances}, vol.~6,
no.~16, eaay2631, 2020.

\bibitem{langley1987scientific}
P.~Langley, H.~A.~Simon, G.~L.~Bradshaw, and J.~M.~Zytkow,
\emph{Scientific Discovery: Computational Explorations of the
Creative Processes}. MIT Press, 1987.

\bibitem{cranmer2023interpretable}
M.~Cranmer, ``Interpretable machine learning for science with PySR
and SymbolicRegression.jl,'' \emph{arXiv:2305.01582}, 2023.

\bibitem{batchprobe}
A.~H.~Bond, batch-probe: GPU memory batch-size probing, v0.3.0
(2026). \url{https://pypi.org/project/batch-probe/}

\end{thebibliography}

\end{document}
